---
title: "August_trend_analysis"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
load_temp <- read_csv(here("data", "willamete_temp1.csv"))
load_temp$month <- lubridate::month(load_temp$Date, label = TRUE)
load_temp$year  <- lubridate::year(load_temp$Date)
load_temp$site_no <- as.character(load_temp$site_no)

aug_data <- subset(load_temp, month == "Aug")
aug_data
aug_model <- glm(temp_C ~ year * site_no, data = aug_data)
summary(aug_model)
```

```{r}
coefs <- coef(aug_model)

site_slopes <- tibble(
  site_no = c("14166000", "14211010", "14211720"),
  slope_per_year = c(
    coefs["year"],
    coefs["year"] + coefs["year:site_no14211010"],
    coefs["year"] + coefs["year:site_no14211720"]
  )
)

site_slopes

aug_data |>
  group_by(site_no) |>
  summarize(
    mean_temp = mean(temp_C, na.rm = TRUE),
    min_temp  = min(temp_C, na.rm = TRUE),
    max_temp  = max(temp_C, na.rm = TRUE),
    n_years   = n_distinct(year)
  )

ggplot(aug_data, aes(x = year, y = temp_C, color = site_no)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "August temperature trends by site, 2010-2023",
    x = "Year",
    y = "Temperature (°C)",
    color = "Site"
  ) +
  theme_minimal()

```

------------------------------------------------------------------------

Conclusions:

1.  Yes, there is a significant trend (**p \< 0.05**) between year and August water temperatures at the reference site (**14166000**), indicating a strong trend in August water temperature over **2010–2023**.
2.  There is a positive trend between water temperature and time. At the reference site (**14166000**), temperatures rose at approximately **0.13** degrees Celsius per year. Over the full 14-year period, that is approximately **1.8** degrees Celsius of warming in August.
3.  The baseline temperatures (mean) differ across sites. The reference site (**14166000**) has the lowest temperature of **18.6** degrees Celsius, site **14211010** has a temperature of **19.9** degrees Celsius, site **14211720** has the highest temperature of **22.8** degrees Celsius. The rate of warming (slope per year) differs across sites. The reference site (**14166000**) has the lowest rate of **0.13** degrees Celsius/year and site **14211720** has a moderate rate of **0.15** degrees Celsius/year. Site **14211010** has the highest rate of **0.18** degrees Celsius/year and is statistically significant from the reference site **14166000**, with a p-value of **0.0097** (p \< **0.05**).
