---
title: "July_trend_analysis"
format: html
editor: visual
---

## 

```{r}
library(tidyverse)
library(here)

load_temp <- read_csv(here("data", "willamete_temp1.csv"))

load_temp$month <- lubridate::month(load_temp$Date, label = TRUE)
load_temp$year  <- lubridate::year(load_temp$Date)
load_temp$site_no  <- as.character(load_temp$site_no
                                   )

july_data <- subset(load_temp, month == "Jul")

july_model <- glm(temp_C ~ year * site_no, data = july_data)
summary(july_model)
```

```{r}
coefs <- coef(july_model)

site_slopes <- tibble(
  site_no = c("14166000", "14211010", "14211720"),
  slope_per_year = c(
    coefs["year"],
    coefs["year"] + coefs["year:site_no14211010"],
    coefs["year"] + coefs["year:site_no14211720"]
  )
)

site_slopes

july_data %>% 
  group_by(site_no) %>% 
  summarize(
    mean_temp = mean(temp_C, na.rm = TRUE),
    min_temp  = min(temp_C, na.rm = TRUE),
    max_temp  = max(temp_C, na.rm = TRUE),
    n_years   = n_distinct(year)
  )


ggplot(july_data, aes(x = year, y = temp_C, color = site_no)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "July temperature trends by site, 2010-2023",
    x = "Year",
    y = "Temperature (°C)",
    color = "Site"
  ) +
  theme_minimal()

```

1.  Yes, there is a significant trend (p < **0.05**) between year and July water temperatures at the reference site (**14166000**), indicating a strong trend in July water temperature over **2010–2023**.

2.  There is a positive trend between water temperature and time. At the reference site (**14166000**), temperatures rose at approximately **0.15** degrees Celsius per year. Over the full 14-year period, that is approximately 2.1 degrees Celsius of warming in July.

3.  The baseline temperatures (mean) differ across sites. The reference site (**14166000**) has the lowest temperature of **18.5** degrees Celsius, site **14211010** has a temperature of **19.5** degrees Celsius, site **14211720** has the highest temperature of **22.1** degrees Celsius. The rate of warming (slope per year) differs across sites. The reference site (**14166000**) has the lowest rate of **0.15** degrees Celsius/year and site **14211720** has a moderate rate of **0.19** degrees Celsius/year. Site **14211010** has the highest rate of **0.23** degrees Celsius/year and and is statistically significant from the reference site **14166000**, with a p-value of **0.002** (p < **0.05**).